---
title: "UPMC Sepsis cleaning"
output: html_document
date: "2024-01-05"
---

```{r setup, include=FALSE}

library(data.table)
library(dplyr)
library(tidyr)
library(skimr)
library(VIM)
library(mice)
library(ggplot2)

upmc <- fread("C:/Users/vadim/OneDrive/Desktop/UPMC/UPMC Emergency cohort pre-imputation carryforward V1.csv", 
                 colClasses = c(gender = "factor", icu = "factor", mechvent = "factor", race = "factor", dead = "factor",
                                surg = "factor",  pbc = "factor", pos_pbc = "factor", cx_type = "factor", 
                                cx_source = "factor", admit_year = "factor", stringsAsFactors = TRUE))

upmc2 <- fread("C:/Users/vadim/OneDrive/Desktop/UPMC/UPMC Emergency cohort original V1.csv", 
               colClasses = c(gender = "factor", icu = "factor", mechvent = "factor", race = "factor", dead = "factor",
                              surg = "factor",  pbc = "factor", pos_pbc = "factor", cx_type = "factor", 
                              cx_source = "factor", admit_year = "factor"))

upmc2 <- subset(upmc2, select = -c(sofa_total, sirs_total, age, alb, alt, ast, base_excess, bicarb, bili, bun, cl, creat, dbp,
                                   elix, fio2, gcs, gender, gluc, hr, hgb, icu, inr, lactate, map, mechvent, paco2, pao2, 
                                   pf_ratio, ph, plt, k, rr, shock_index, na, o2_sat, sbp, temp, wbc, weight, fluid, 
                                   norepi_equiv, dtt0_hours, dead_90, action, interval, empi_id))

heights <- fread("C:/Users/vadim/OneDrive/Desktop/UPMC/UPMC Heights.csv")

```

## Join datasets
merge by hosp_id and row number (id)

Saved as:
write.csv(cupmc, "C:/Users/vadim/OneDrive/Desktop/UPMC/UPMC Combined long data.csv")

```{r merge}

upmc <- group_by(upmc, hosp_id) |> mutate(id = row_number())
upmc2 <- group_by(upmc2, hosp_id) |> mutate(id = row_number())

mupmc <- full_join(upmc, upmc2, by = c("hosp_id", "id"))

heights <- group_by(heights, hosp_id) |> mutate(id = row_number())

cupmc <- right_join(heights, mupmc, by = c("hosp_id", "id"))

setorder(cupmc, hosp_id, id)

```


## OPTIONAL:  Limiting dataset to sepsis-associated hypotension or shock
length(unique(fupmc$hosp_id))
N = 43462

table(fupmc$id)
 1     2     3     4     5     6     7     8     9    10    11    12    13 
 43462 43462 43078 42872 42603 42254 41776 41309 40951 40627 40082 39260 13601

```{r Hypotension or Shock}
fupmc <- cupmc |>
  group_by(hosp_id) |>                
  filter(any(sbp < 101 | map < 66 | norepi_equiv > 0)) |>  
  ungroup()   

```

## Making a new time variable for t=0 == 1st time someone is recorded as being HoTN or in shock
May or may not use in analysis

saved as: 
write.csv(fupmc, "C:/Users/vadim/OneDrive/Desktop/UPMC/UPMC combined_pre imputed.csv")

```{r}

create_time_column <- function(fupmc) {
  fupmc |>
    group_by(hosp_id) |>
    mutate(first_row = which.min(ifelse(sbp < 101 | map < 66 | norepi_equiv > 0, interval, Inf))) |>
    mutate(time = row_number() - first_row) |>
    ungroup() 
}

fupmc <- create_time_column(fupmc)

fupmc |> relocate(time)
```

## Carryforward heights
when not missing

sum(is.na(cupmc$height))
= 22959

can also carry forward with library(zoo)
tmleupmc$hospitallocf <- zoo::na.locf(tmleupmc$hospital)
sum(is.na(tmleupmc$hospitallocf))

```{r carryforward heights}

cupmc <- cupmc |> 
  group_by(hosp_id) |>
  fill(height, .direction = "down") |>
  ungroup()

```

## Hot deck imputation
Not enough compute for other imputations


#Here's code for KNN:
knn_fupmc <- fupmc[, !(names(fupmc) %in% c("fluid_cckg_ibw", "adjbw", 
                                          "ibw", "fluid_cckg", "egfr", "pulsepress", "pf_ratio", "shock_index"))]

numeric_fupmc <- fupmc[sapply(fupmc, is.numeric)]
nonnumeric_fupmc <- fupmc[sapply(fupmc, Negate(is.numeric))]

knn_normupmc <- as.data.frame(lapply(numeric_fupmc, scale))

knn_fupmc <- fupmc[, !(names(fupmc) %in% c("pf_ratio", "shock_index", "enc_start_dt", "enc_end_dt", "dt_t0_first",  "death_date",
                                          "icu_start_dt", "icu_end_dt", "hospital", "empi_id"))]

knn_fupmc <- knn_normupmc[, !(names(knn_normupmc) %in% c("pf_ratio", "shock_index"))]

knn_upmc5 <- kNN(knn_fupmc, k = 5)

#And code for MICE (which I can't get to converge):
imp_mice <- mice(fupmc[, c("height", "weight", "alb", "alt", "ast", "base_excess", "elix",        
                   "bicarb", "bili", "bun", "cl", "creat", "dbp", "fio2", "gcs", 
                   "gluc", "hr", "hgb", "inr", "lactate", "map", "paco2", "pao2",                 
                   "pf_ratio", "ph", "plt", "k", "rr", "na", "o2_sat","sbp", "temp",                 
                   "wbc")], 
            method = 'pmm', 
            m = 5, 
            maxit = 15,
            seed = 1,
            printFlag = TRUE)

compl_fupmc <- complete(imp_mice, action = "long", include = TRUE)

summary(compl_fupmc)

summary(fupmc)

head(compl_upmc[compl_upmc$.imp_mice == 2, ], )

--diagnosing imputation problems: densityplot(imp_mice), stripplot(imp_mice)
sapply(fupmc, function(i) sum(is.na(i)))

```{r hot deck imputation}
hdupmc <- hotdeck(cupmc)

sum(is.na(fupmc_impuhd))

```

## Defining new clinical variables
--pulse pressure
--eGFR
--IBW
--Adjusted BW
--fluid volumes / actual, adjusted, and ideal body weights
--Did not break up SOFA into component scores

saved as: 
write.csv(hdupmc, "C:/Users/vadim/OneDrive/Desktop/UPMC/UPMC hd_imputed new_vars all_pts no_time.csv")

```{r new variables}
#pulse pressure
hdupmc <- hdupmc |>
  mutate(pulsepress = sbp - dbp)

#eGFR
calculate_egfr <- function(creat, age, gender) {
  a <- ifelse(gender == "0", 101.9, 138.2)
  b <- ifelse(gender == "0", 0.80, 1.05)
  c <- ifelse(gender == "0", -0.742, -0.918)
  
  egfr <- a * ((creat / b) ^ c) * (0.993 ^ age)
  
  return(egfr)
}

hdupmc <- hdupmc |>
  mutate(egfr = calculate_egfr(creat, age, gender))


#fluid / actual weight
hdupmc <- hdupmc |>
  mutate(fluid_cckg = fluid_sum / weight)

#fluid / adjusted weight
hdupmc$adjbw <- hdupmc$ibw + 0.4 * (hdupmc$weight - hdupmc$ibw)

hdupmc <- hdupmc |>
  mutate(fluid_cckg_abw = fluid_sum / adjbw)

#fluid / ideal weight
hdupmc <- hdupmc |>
         mutate(ibw = pmax(0.1, if_else(gender == 1,
                               50 + 0.91 * (height - 152.4),
                               45.5 + 0.91 * (height - 152.4))))

hdupmc <- hdupmc |>
  mutate(fluid_sum_cckgibw = fluid_sum / ibw)


write.csv(hdupmc, "C:/Users/vadim/OneDrive/Desktop/UPMC/UPMC hd_imputed new_vars all_pts no_time.csv")

```

##Confirm no missingness
```{r}
sapply(hdupmc, function(i) sum(is.na(i)))
```


## Skim

```{r skim the data, echo=FALSE}
skim(hdupmc)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
